Perturbing SU($n$)-symmetric interactions

### 2020-02-11 collective_sun/sun_perturbation.tex
added scalar component of operator products for completion / debugging purposes

Here we simplify the product
\begin{multline}
  \P_0 \sp{\sum_{k<\ell} \sum_{p<q}
    v_{k\ell} w_{pq} O_{k\ell} Q_{pq}} \P_0 \\
  \EQPS \diagram{two_body_0} O_{1,2} Q_{3,4}
  + \diagram{two_body_1} O_{1,2} Q_{1,3}
  + \diagram{two_body_2} O_{1,2} Q_{1,2},
  \label{eq:POQP_start}
\end{multline}
where $\P_0$ is a projector onto the permutationally symmetric
manifold $\M_0$; $k,\ell,p,q$ index individual spins;
$v_{k\ell},w_{pq}$ are scalar coefficients with $m_{pq}=m_{qp}$ and
$m_{pp}=0$ for each of $m\in\set{v,w}$; and $O,Q$ are
permutationally-symmetric two-spin operators.  This product appears,
for example, in the calculation of the second-order effective
Hamiltonian $H_2^{(2)}$ induced on the permutationally symmetric
manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} in \eqref{eq:perturbations}.  Defining
\begin{align}
  \v{\m m} \equiv \sum_{p<q} m_{pq} \ket{\set{p,q}},
  &&
  \v m \equiv \sum_p m_p \ket{p},
  &&
  m_p \equiv \sum_q m_{pq},
  &&
  \col{m} \equiv \sum_{p<q} m_{pq},
\end{align}
for $m\in\set{v,w}$, we can write
\begin{align}
  \diagram{two_body_2} \equiv \sum_{\p{p,q}\in\C_N\p{2}} v_{pq} w_{pq}
  = \v{\m v} \c\v{\m w},
\end{align}
and simplify
\begin{align}
  \diagram{two_body_1}
  &= \sp{\diagram{two_body_1_o}} - \sp{\diagram{two_body_1_x}} \\
  &= \sp{\diagram{two_body_1_oo} - \diagram{two_body_1_ox}}
  - \sp{2\diagram{two_body_2}} \label{eq:mid_zero} \\
  &= \v v \c \v w - 2 \v{\m v} \c \v{\m w},
\end{align}
where the middle diagram in \eqref{eq:mid_zero} contains an empty sum
for the cross, and is therefore equal to zero.  Finally, with a bit of
more work (or a computer) we can work out that
\begin{align}
  \diagram{two_body_0}
  = \diagram{two_body_0_oooo} - \diagram{two_body_1_ooo}
  + \diagram{two_body_2_oo}
  = \col{v}\,\col{w} -\v v\c\v w + \v{\m v}\c\v{\m w}.
\end{align}
Altogether,
\begin{multline}
  \sum_{k<\ell} \sum_{p<q} v_{k\ell} w_{pq} O_{k\ell} Q_{pq} \\
  \EQPS \p{\col{v}\,\col{w} -\v v\c\v w + \v{\m v}\c\v{\m w}}
  O_{1,2} Q_{3,4}
  + \p{\v v\c\v w - 2 \v{\m v} \c \v{\m w}} O_{1,2} Q_{1,3}
  + \v{\m v} \c \v{\m w}\, O_{1,2} Q_{1,2}.
  \label{eq:POQP}
\end{multline}
In the special case that $O=Q=X\otimes X$ with $X^2=1$, we can further
simplify
\begin{align}
  \sum_{k<\ell} \sum_{p<q} v_{k\ell} w_{pq} X_k X_\ell X_p X_q
  \simeq_\PS
  \p{\col{v}\,\col{w} -\v v\c\v w + \v{\m v}\c\v{\m w}} X^{\otimes 4}
  + \p{\v v\c\v w - 2 \v{\m v} \c \v{\m w}} X^{\otimes 2},
\end{align}
where $\simeq_\PS$ denotes equality up to scalar terms and a
restriction to the permutationally symmetric manifold.  In terms of
collective operators $\col{X} \equiv \sum_p X_p$, we then have
\begin{multline}
  \sum_{k<\ell} \sum_{p<q} v_{k\ell} w_{pq} X_k X_\ell X_p X_q \\
  \simeq_\PS
  \f{\col{v}\,\col{w} -\v v\c\v w + \v{\m v}\c\v{\m w}}{N!_4}
  \sp{\col{X}^4 - 2\p{3N-4} \col{X}^2}
  + \f{\p{\v v\c\v w - 2 \v{\m v} \c \v{\m w}}}{N!_2}\, \col{X}^2,
\end{multline}
where
\begin{align}
  N!_k \equiv \prod_{j=0}^{k-1} \p{N-j} = \f{N!}{\p{N-k}!}
\end{align}
denotes a falling factorial.

@@

Here we simplify the product
\begin{align}
  \X \equiv \P_0 \sp{\sum_{k<\ell} \sum_{p<q} \sum_{r<s}
    u_{k\ell} v_{pq} w_{rs} X_k X_\ell X_p X_q X_r X_s} \P_0,
  &&
  X^2 = 1,
\end{align}
where $\P_0$ is a projector onto the permutationally symmetric
manifold $\M_0$; $k,\ell,p,q,r,s$ index individual spins; $X$ is a
single-spin operator; and $u_{k\ell},v_{pq},w_{rs}$ are scalar
coefficients with $m_{pq}=m_{qp}$ and $m_{pp}=0$ for each of
$m\in\set{u,v,w}$.  Collecting terms according to the number of $X$
operators that remain after accounting for $X^2=1$, we find that
\begin{align}
  \X \EQPS A_6 X^{\otimes 6} + A_4 X^{\otimes 4}
  + A_2 X^{\otimes 2} + A_0,
  \label{eq:triple_multi}
\end{align}
with
\begin{align}
  A_6 \equiv \diagram{triple_0},
  &&
  A_4 \equiv \diagram{triple_01} + \diagram{triple_1},
  &&
  A_0 \equiv \diagram{triple_0111},
  \label{eq:triple_64}
\end{align}
\begin{align}
  A_2 \equiv \diagram{triple_011} + \diagram{triple_02}
  + \diagram{triple_11} + \diagram{triple_2},
  \label{eq:triple_2}
\end{align}
where a diagram with unlabeled circles denotes a sum over all distinct
label assignments, e.g.
\begin{align}
  \diagram{triple_1} \equiv \diagram{triple_1_uvw},
  \label{eq:triple_uvw}
\end{align}
\begin{align}
  \diagram{triple_011}
  \equiv \diagram{triple_011_uvw}
  + \diagram{triple_011_vwu} + \diagram{triple_011_wuv}.
  \label{eq:triple_uvw_3}
\end{align}
The first diagram in each of $A_6,A_4,A_2$ as written in
\eqref{eq:triple_64}, \eqref{eq:triple_2} thus has only one assignment
of labels, as in \eqref{eq:triple_uvw}, while all other unlabeled
diagrams have three assignments, as in \eqref{eq:triple_uvw_3}.

Having identified diagrammatic representations of the coefficients
$A_6,A_4,A_2$, we now need to simplify all diagrams to reduce their
computational complexity, as discussed in Section \ref{sec:diagrams}.
Simplifying all these diagrams by hand is tedious, but the process of
eliminating filled dots and crosses from a diagram is algorithmic
enough to execute on a computer, doing which we find
\begin{align}
  A_6 = D_0 - D_1 + D_2 - D_3,
  &&
  A_4 = D_1 - 2 D_2 + 3 D_3,
  &&
  A_2 = D_2 - 3 D_3,
  &&
  A_0 \equiv D_3,
\end{align}
where
\begin{align}
  D_0 \equiv \diagram{triple_0_o},
  &&
  D_1 \equiv \diagram{triple_01_o} - 2 \diagram{triple_1_o},
  &&
  D_3 \equiv \diagram{triple_0111_o},
\end{align}
\begin{align}
  D_2 \equiv \diagram{triple_011_o}
  + \diagram{triple_02_o}
  - 2 \diagram{triple_11_o}
  + 4 \diagram{triple_2_o}.
\end{align}
Defining, for each of $m\in\set{u,v,w}$,
\begin{align}
  \m m \equiv \sum_{p,q} m_{pq} \op{p}{q},
  &&
  \v{\m m} \equiv \sum_{p<q} m_{pq} \ket{\set{p,q}},
  &&
  \v m \equiv \sum_p m_p \ket{p},
  &&
  m_p \equiv \sum_q m_{pq},
  &&
  \col{m} \equiv \sum_{p<q} m_{pq},
\end{align}
we can expand
\begin{align}
  D_0 = \col{u}\,\col{v}\,\col{w},
  &&
  D_1 = \col{u}\,\v v\c\v w + \col{v}\,\v w\c\v u
  + \col{w}\,\v u\c\v v - 2 \sum_p u_p v_p w_p,
  &&
  D_3 = \tr\p{\m u \c \m v \c \m w},
\end{align}
\begin{multline}
  D_2 = \v u \c\m v\c\v w + \v v \c\m w\c\v u + \v w \c\m u\c\v v
  + \col{u}\,\v{\m v}\c\v{\m w} + \col{v}\,\v{\m w}\c\v{\m u}
  + \col{w}\,\v{\m u}\c\v{\m v} \\
  - 2 \sum_{p,q} \p{u_p v_{pq} w_{pq}
    + v_p w_{pq} u_{pq} + w_p u_{pq} v_{pq}}
  + 4 \sum_{p<q} u_{pq} v_{pq} w_{pq}.
\end{multline}
Finally, we can also write the product $\X$ in \eqref{eq:triple_multi}
in terms of the collective operator $\col{X} \equiv \sum_p X_p$:
\begin{align}
  \X \simeq_\PS
  \tilde A_6 \col{X}^6 + \tilde A_4 \col{X}^4 + \tilde A_2 \col{X}^2,
  \label{eq:triple_col}
\end{align}
where $\simeq_\PS$ denotes equality up to scalar terms and a
restriction to the permutationally symmetric manifold, and
\begin{align}
  \tilde A_6 \equiv \f{A_6}{N!_6},
  &&
  \tilde A_4 \equiv \f{A_4}{N!_4} - 5\p{3N-8} \f{A_6}{N!_6},
\end{align}
\begin{align}
  \tilde A_2 \equiv \f{A_2}{N!_2} - 2\p{3N-4} \f{A_4}{N!_4}
  + \p{45N^2-210N+184} \f{A_6}{N!_6}.
\end{align}

>>>>

Here we simplify the product
\begin{multline}
  \P_0 \sp{\sum_{k<\ell} \sum_{p<q}
    v_{k\ell} w_{pq} O_{k\ell} Q_{pq}} \P_0 \\
  \EQPS \diagram{two_body_0} O_{1,2} Q_{3,4}
  + \diagram{two_body_1} O_{1,2} Q_{1,3}
  + \diagram{two_body_2} O_{1,2} Q_{1,2},
  \label{eq:POQP_start}
\end{multline}
where $\P_0$ is a projector onto the permutationally symmetric
manifold $\M_0$; $k,\ell,p,q$ index individual spins;
$v_{k\ell},w_{pq}$ are scalar coefficients with $m_{pq}=m_{qp}$ and
$m_{pp}=0$ for each of $m\in\set{v,w}$; and $O,Q$ are
permutationally-symmetric two-spin operators.  This product appears,
for example, in the calculation of the second-order effective
Hamiltonian $H_2^{(2)}$ induced on the permutationally symmetric
manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} in \eqref{eq:perturbations}.  Defining
\begin{align}
  \v{\m m} \equiv \sum_{p<q} m_{pq} \ket{\set{p,q}},
  &&
  \v m \equiv \sum_p m_p \ket{p},
  &&
  m_p \equiv \sum_q m_{pq},
  &&
  \col{m} \equiv \sum_{p<q} m_{pq},
\end{align}
for $m\in\set{v,w}$, we can write
\begin{align}
  \diagram{two_body_2} \equiv \sum_{\p{p,q}\in\C_N\p{2}} v_{pq} w_{pq}
  = \v{\m v} \c\v{\m w},
\end{align}
and simplify
\begin{align}
  \diagram{two_body_1}
  &= \sp{\diagram{two_body_1_o}} - \sp{\diagram{two_body_1_x}} \\
  &= \sp{\diagram{two_body_1_oo} - \diagram{two_body_1_ox}}
  - \sp{2\diagram{two_body_2}} \label{eq:mid_zero} \\
  &= \v v \c \v w - 2 \v{\m v} \c \v{\m w},
\end{align}
where the middle diagram in \eqref{eq:mid_zero} contains an empty sum
for the cross, and is therefore equal to zero.  Finally, with a bit of
more work (or a computer) we can work out that
\begin{align}
  \diagram{two_body_0}
  = \diagram{two_body_0_oooo} - \diagram{two_body_1_ooo}
  + \diagram{two_body_2_oo}
  = \col{v}\,\col{w} -\v v\c\v w + \v{\m v}\c\v{\m w}.
\end{align}
Altogether,
\begin{multline}
  \sum_{k<\ell} \sum_{p<q} v_{k\ell} w_{pq} O_{k\ell} Q_{pq} \\
  \EQPS \p{\col{v}\,\col{w} -\v v\c\v w + \v{\m v}\c\v{\m w}}
  O_{1,2} Q_{3,4}
  + \p{\v v\c\v w - 2 \v{\m v} \c \v{\m w}} O_{1,2} Q_{1,3}
  + \v{\m v} \c \v{\m w}\, O_{1,2} Q_{1,2}.
  \label{eq:POQP}
\end{multline}
In the special case that $O=Q=X\otimes X$ with $X^2=1$, we can further
simplify
\begin{align}
  \sum_{k<\ell} \sum_{p<q} v_{k\ell} w_{pq} X_k X_\ell X_p X_q
  \EQPS
  \p{\col{v}\,\col{w} -\v v\c\v w + \v{\m v}\c\v{\m w}} X^{\otimes 4}
  + \p{\v v\c\v w - 2 \v{\m v} \c \v{\m w}} X^{\otimes 2}
  + \v{\m v}\c\v{\m w}.
\end{align}
In terms of collective operators $\col{X} \equiv \sum_p X_p$, we then
have
\begin{multline}
  \sum_{k<\ell} \sum_{p<q} v_{k\ell} w_{pq} X_k X_\ell X_p X_q \\
  \EQPS
  \f{\col{v}\,\col{w} -\v v\c\v w + \v{\m v}\c\v{\m w}}{N!_4}
  \sp{\col{X}^4 - 2\p{3N-4} \col{X}^2}
  + \f{\p{\v v\c\v w - 2 \v{\m v} \c \v{\m w}}}{N!_2}\, \col{X}^2 \\
  + \f{3\,\col{v}\,\col{w} - N \v v\c\v w
    + N\p{N-2}\v{\m v}\c\v{\m w}}{\p{N-1}\p{N-3}},
\end{multline}
where
\begin{align}
  N!_k \equiv \prod_{j=0}^{k-1} \p{N-j} = \f{N!}{\p{N-k}!}
\end{align}
denotes a falling factorial.

@@

Here we simplify the product
\begin{align}
  \X \equiv \P_0 \sp{\sum_{k<\ell} \sum_{p<q} \sum_{r<s}
    u_{k\ell} v_{pq} w_{rs} X_k X_\ell X_p X_q X_r X_s} \P_0,
  &&
  X^2 = 1,
\end{align}
where $\P_0$ is a projector onto the permutationally symmetric
manifold $\M_0$; $k,\ell,p,q,r,s$ index individual spins; $X$ is a
single-spin operator; and $u_{k\ell},v_{pq},w_{rs}$ are scalar
coefficients with $m_{pq}=m_{qp}$ and $m_{pp}=0$ for each of
$m\in\set{u,v,w}$.  Collecting terms according to the number of $X$
operators that remain after accounting for $X^2=1$, we find that
\begin{align}
  \X \EQPS A_6 X^{\otimes 6} + A_4 X^{\otimes 4}
  + A_2 X^{\otimes 2} + A_0,
  \label{eq:triple_multi}
\end{align}
with
\begin{align}
  A_6 \equiv \diagram{triple_0},
  &&
  A_4 \equiv \diagram{triple_01} + \diagram{triple_1},
  &&
  A_0 \equiv \diagram{triple_0111},
  \label{eq:triple_64}
\end{align}
\begin{align}
  A_2 \equiv \diagram{triple_011} + \diagram{triple_02}
  + \diagram{triple_11} + \diagram{triple_2},
  \label{eq:triple_2}
\end{align}
where a diagram with unlabeled circles denotes a sum over all distinct
label assignments, e.g.
\begin{align}
  \diagram{triple_1} \equiv \diagram{triple_1_uvw},
  \label{eq:triple_uvw}
\end{align}
\begin{align}
  \diagram{triple_011}
  \equiv \diagram{triple_011_uvw}
  + \diagram{triple_011_vwu} + \diagram{triple_011_wuv}.
  \label{eq:triple_uvw_3}
\end{align}
The first diagram in each of $A_6,A_4,A_2,A_0$ as written in
\eqref{eq:triple_64}, \eqref{eq:triple_2} thus has only one assignment
of labels, as in \eqref{eq:triple_uvw}, while all other unlabeled
diagrams have three assignments, as in \eqref{eq:triple_uvw_3}.

Having identified diagrammatic representations of the coefficients
$A_6,A_4,A_2,A_0$, we now need to simplify all diagrams to reduce
their computational complexity, as discussed in Section
\ref{sec:diagrams}.  Simplifying all these diagrams by hand is
tedious, but the process of eliminating filled dots and crosses from a
diagram is algorithmic enough to execute on a computer, doing which we
find
\begin{align}
  A_6 = D_0 - D_1 + D_2 - D_3,
  &&
  A_4 = D_1 - 2 D_2 + 3 D_3,
  &&
  A_2 = D_2 - 3 D_3,
  &&
  A_0 \equiv D_3,
\end{align}
where
\begin{align}
  D_0 \equiv \diagram{triple_0_o},
  &&
  D_1 \equiv \diagram{triple_01_o} - 2 \diagram{triple_1_o},
  &&
  D_3 \equiv \diagram{triple_0111_o},
\end{align}
\begin{align}
  D_2 \equiv \diagram{triple_011_o}
  + \diagram{triple_02_o}
  - 2 \diagram{triple_11_o}
  + 4 \diagram{triple_2_o}.
\end{align}
Defining, for each of $m\in\set{u,v,w}$,
\begin{align}
  \m m \equiv \sum_{p,q} m_{pq} \op{p}{q},
  &&
  \v{\m m} \equiv \sum_{p<q} m_{pq} \ket{\set{p,q}},
  &&
  \v m \equiv \sum_p m_p \ket{p},
  &&
  m_p \equiv \sum_q m_{pq},
  &&
  \col{m} \equiv \sum_{p<q} m_{pq},
\end{align}
we can expand
\begin{align}
  D_0 = \col{u}\,\col{v}\,\col{w},
  &&
  D_1 = \col{u}\,\v v\c\v w + \col{v}\,\v w\c\v u
  + \col{w}\,\v u\c\v v - 2 \sum_p u_p v_p w_p,
  &&
  D_3 = \tr\p{\m u \c \m v \c \m w},
\end{align}
\begin{multline}
  D_2 = \v u \c\m v\c\v w + \v v \c\m w\c\v u + \v w \c\m u\c\v v
  + \col{u}\,\v{\m v}\c\v{\m w} + \col{v}\,\v{\m w}\c\v{\m u}
  + \col{w}\,\v{\m u}\c\v{\m v} \\
  - 2 \sum_{p,q} \p{u_p v_{pq} w_{pq}
    + v_p w_{pq} u_{pq} + w_p u_{pq} v_{pq}}
  + 4 \sum_{p<q} u_{pq} v_{pq} w_{pq}.
\end{multline}
Finally, we can also write the product $\X$ in \eqref{eq:triple_multi}
in terms of the collective operator $\col{X} \equiv \sum_p X_p$:
\begin{align}
  \X \EQPS
  \tilde A_6 \col{X}^6 + \tilde A_4 \col{X}^4
  + \tilde A_2 \col{X}^2 + \tilde A_0,
  \label{eq:triple_col}
\end{align}
where
\begin{align}
  \tilde A_6 \equiv \f{A_6}{N!_6},
  &&
  \tilde A_4 \equiv \f{A_4}{N!_4} - 5\p{3N-8} \f{A_6}{N!_6},
\end{align}
\begin{align}
  \tilde A_2 \equiv \f{A_2}{N!_2} - 2\p{3N-4} \f{A_4}{N!_4}
  + \p{45N^2-210N+184} \f{A_6}{N!_6},
\end{align}
\begin{align}
  \tilde A_0 \equiv A_0 - \f{A_2}{N-1}
  + \f{3A_4}{\p{N-1}\p{N-3}}
  - \f{15A_6}{\p{N-1}\p{N-3}\p{N-5}}.
\end{align}
